Density-dependent selection and the population ecology of relative fitness

### 2017-12-01 manuscript.tex
l

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. This variety is not necessarily problematic in itself, because different measures of fitness may be more useful in different circumstances. But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While this grounding is clear for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.

In uncrowded populations, relative fitness simply represents differences in the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}, with selection favoring greater $r$ ($r$-selection). In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course. This is exemplified by the Wright-Fisher model, in which time advances in discrete non-overlapping generations, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. The limitations of these relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating more of the life cycle ``components'' of fitness (e.g. \cite[pp. 276]{ewens_2004}). But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.

The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Apart from the type densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}. The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction. Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.

@@

Constant-$N$ relative fitness models are not compatible with $K$-selection, and must refer to the competitive effects summarized by $K_{12}$ and $K_{21}$. We thus now restrict our attention to the case $K_{11}=K_{22}$. Fig.~\ref{fig:LVvslottery} shows two examples of selection between two types with the same saturation density. Fig.~\ref{fig:LVvslottery}a shows exclusion by a superior Lotka-Volterra competitor ($\alpha_{21}>\alpha_{12}$). Even though the saturation densities are identical, the density trajectories do not follow a line of constant $N$. This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over a selective sweep.

>>>>

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. This variety is not necessarily problematic in itself, because different measures of fitness may be more useful in different circumstances. But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While this grounding is clear for absolute fitness measures like $r$ and saturation density, relative fitness seems largely divorced from population ecology.

In uncrowded populations, relative fitness simply represents differences in the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}, with selection favoring greater $r$ ($r$-selection). The situation is considerably more complicated in crowded populations. In this case, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course. This is exemplified by the Wright-Fisher model, in which time advances in discrete non-overlapping generations, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. The limitations of relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating multiple life cycle ``components'' of fitness (e.g. \cite[pp. 276]{ewens_2004} \citep{prout_1980}). But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.

The issue can be expressed formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Apart from the type densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}. The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the broader gulf with population ecology \citep{mallet_2012}. Thus, it is important that we deepen our population-ecological understanding of relative fitness models, both to gain insight into their domain of applicability, and as part of the broad challenge of synthesizing ecology and evolution.

The recent rise of adaptive dynamics has provided a powerful framework for modeling the interplay between long-term evolution and population ecology, including complex couplings between evolutionary change and population density. However, the focus of this latter work is on trait evolution rather than the underlying genetics, and in particular, the details of selective sweeps are typically subsumed into effectively instantaneous ``trait substitutions'' determined by invasion fitness. We emphasize that our focus here is on the description of selection itself, which is critical for evolutionary genetics i.e. for making sense of the change in allele frequencies over time.

@@

Constant-$N$ relative fitness models are not compatible with $K$-selection, and must refer to the competitive effects summarized by $K_{12}$ and $K_{21}$. We thus now restrict our attention to the case $K_{11}=K_{22}$. Fig.~\ref{fig:LVvslottery} shows two examples of selection between two types with the same saturation density. Fig.~\ref{fig:LVvslottery}a shows exclusion by a superior Lotka-Volterra competitor ($\alpha_{21}>\alpha_{12}$). Even though the saturation densities are identical, the density trajectories do not follow a line of constant $N$. This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ due to the selective sweep.
